Local PCA Shows How the Effect of Population Structure Differs Along the Genome

### 2016-09-07 writeup/local_pca_paper.tex
poster

To help visualize how clustered windows with similar patterns of relatedness are along each chromosome arm, 
we selected three ``extreme'' windows in the MDS plot
and the 5\% of windows that are closest to it in the MDS coordinates,
then highlighted these windows' positions along the genome,
and created PCA plots for the windows, combined.
Representative plots are shown for three groups of windows on each chromosome arm
in Figure \ref{fig:mds_allchr} (groups are shown in color),
and in Figure \ref{fig:pca_by_pop} (PCA plots).
The latter plots are quite different,
showing that genomic windows in different regions of the MDS plot indeed show quite different patterns of relatedness:

@@

Unlike the other two species,
the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
showed similar patterns of gradual change in patterns of relatedness across each chromosome,
with no indications of chromosome-specific patterns.
This consistency suggests that the factor affecting the population structure for each chromosome is the same,
as might be caused by varying strengths of linked selection.
To verify that variation in the effects of population structure is shared across chromosomes,
we applied the method to all chromosomes together.
Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
and other chromosomes are similar:
% the two types of population structure with higher gene density 
% (colored green and purple in Figure \ref{fig:mds12_medicago})
% are nearly identical across all chromosomes,
% but population structure associated with low gene density (colored orange)
% shows more chromsome-specific patterns (see Supplemental Figure \ref{sfig:pca_drosophila_allchr}).
across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
which often have lower gene density and may therefore be less subject to linked selection.
% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
To verify that this is a possible explanation,
we computed gene densities in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
which are shown juxtaposed with
the first MDS coordinate of each window in Figure \ref{fig:mds_medicago},
and are significantly correlated, as shown in Supplemental Figure \ref{sfig:mds_gene_count}.
(Values shown are the number of start and end positions of each predicted mRNA transcript,
divided by two, assigned to the nearest window.)
% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
However, other genomic features, such as distance to centromere show roughly the same patterns,
so we cannot rule out alternative hypotheses.
In particular, the recombination rates estimated by \citet{paape2012finescale} appear visually to be similar,
but were not readily available for comparison.

@@

\begin{figure}
    \begin{center}
       \includegraphics{Fig6_Together_MDS_plot_chr3_final}
    \end{center}
    \caption{
       MDS visualization of patterns of relatedness on \textit{M.~truncatula} chromosome 3. 
       Each point in the plot represents a window;
       the structure revealed by the MDS plot is strongly clustered along the chromosome,
       with windows in the upper-right corner of the MDS plot (colored red) clustered around the centromere,
       windows in the upper-left corner (purple) furthest from the centromere,
       and the remaining corner (green) intermediate.
       Plots for remaining chromosomes are shown in Supplemental Figure \ref{sfig:mds_medicago_allchr}.
       \label{fig:mds12_medicago}
    }
\end{figure}

\begin{figure}
    \begin{center}
       \includegraphics{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
    \end{center}
    \caption{
        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in Figure \ref{fig:mds12_medicago}. 
        Each point corresponds to a sample, colored by country of origin.
        Plots for remaining chromosomes are shown in Supplemental Figure \ref{sfig:pca_peaks_medicago_allchr}.
        \label{fig:pca_medicago}
    }
\end{figure}

@@

%%%%%%%%
\subsection*{Acknowledgements}

@@

\subsection{Data accessibility}

@@

>>>>

To help visualize how clustered windows with similar patterns of relatedness are along each chromosome arm, 
we selected three ``extreme'' windows in the MDS plot
and the 5\% of windows that are closest to it in the MDS coordinates,
then highlighted these windows' positions along the genome,
and created PCA plots for the windows, combined.
Representative plots are shown for three groups of windows on each chromosome arm
in Figure \ref{fig:mds_allchr} (groups are shown in color),
and in Supplemental Figure \ref{fig:pca_by_pop} (PCA plots).
The latter plots are quite different,
showing that genomic windows in different regions of the MDS plot indeed show quite different patterns of relatedness.

% Fig fig:pca_by_pop WAS HERe

@@

Unlike the other two species,
the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
showed similar patterns of gradual change in patterns of relatedness across each chromosome,
with no indications of chromosome-specific patterns.
This consistency suggests that the factor affecting the population structure for each chromosome is the same,
as might be caused by varying strengths of linked selection.
To verify that variation in the effects of population structure is shared across chromosomes,
we applied the method to all chromosomes together.
Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
and other chromosomes are similar:
% the two types of population structure with higher gene density 
% (colored green and purple in Figure \ref{fig:mds12_medicago})
% are nearly identical across all chromosomes,
% but population structure associated with low gene density (colored orange)
% shows more chromsome-specific patterns (see Supplemental Figure \ref{sfig:pca_drosophila_allchr}).
across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
which often have lower gene density and may therefore be less subject to linked selection.
% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
To verify that this is a possible explanation,
we counted the number of genes found in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
which are shown juxtaposed with
the first MDS coordinate of each window in Figure \ref{fig:mds_medicago},
and are significantly correlated, as shown in Supplemental Figure \ref{sfig:mds_gene_count}.
(Values shown are the number of start and end positions of each predicted mRNA transcript,
divided by two, assigned to the nearest window.)
% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
However, other genomic features, such as distance to centromere show roughly the same patterns,
so we cannot rule out alternative hypotheses.
In particular, the recombination rates estimated by \citet{paape2012finescale} appear visually to be similar,
but were not available in a form that can be compared to our MDS values.

@@

\begin{figure}
    \begin{center}
       \includegraphics{Fig6_Together_MDS_plot_chr3_final}
       \includegraphics{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
    \end{center}
    \caption{
       MDS visualization of patterns of relatedness on \textit{M.~truncatula} chromosome 3,
       with corresponding PCA plots.
       Each point in the plot represents a window;
       the structure revealed by the MDS plot is strongly clustered along the chromosome,
       with windows in the upper-right corner of the MDS plot (colored red) clustered around the centromere,
       windows in the upper-left corner (purple) furthest from the centromere,
       and the remaining corner (green) intermediate.
       Plots for remaining chromosomes are shown in Supplemental Figure \ref{sfig:mds_medicago_allchr}.
       \textbf{(below)}
        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in Figure \ref{fig:mds12_medicago}. 
        Each point corresponds to a sample, colored by country of origin.
        Plots for remaining chromosomes are shown in Supplemental Figure \ref{sfig:pca_peaks_medicago_allchr}.
       \label{fig:mds12_medicago}
       \label{fig:pca_medicago}
    }
\end{figure}

@@

@@

\section{Data access}

@@

Datasets are available as follows:
human (POPRES) at dbGaP with accession number phs000145.v4.p2,
\textit{Medicago} at the Medicago Hapmap \url{http://www.medicagohapmap.org/},
and \textit{Drosophila} at the Drosophila Genome Nexus, \url{http://www.johnpool.net/genomes.html}.

%%%%%%%%
\section*{Acknowledgements}

%%%%%%%%
\section*{Disclosure declaration}

The authors declare no conflicts of interest.

### 2016-09-07 writeup/submission/cover-letter.tex
poster

We are writing to submit our paper,
``Local PCA Shows How the Effect of Population Structure Differs Along the Genome'',
for consideration in \textit{JOURNAL XXX}.
In this paper, we set out to describe how background patterns of relatedness 
(colloquially, ``population structure'')
vary along the genome at a scale larger than correlations induced by linkage.
It was not certain that we would find anything of interest:
in the field we tend to think of neutral, background population structure as a single thing,
against which outlying signals of selection can be identified.
What we found was not only substantial variation at chromosomal scales,
but different underlying causes in the three species we consider.
The focus of our paper is on the method we developed to identify and visualize such variation in patterns of relatedness.
We believe our method represents a fundamental advance because it lets the data speak for itself,
providing an unbiased view of what are the most important variation in patterns of relatedness,
rather than looking for specific signals.
The method is therefore complementary to more common methods that identify small regions of the genome
that are outliers for particular summaries of relatedness (e.g., $F_{ST}$-based scans for selection).
We think the paper will be of interest to readers of \textit{JOURNAL XXX}
because the method fills an important gap in the genomicist's toolkit,
that is tailored to describing variation in modern, spatially sampled datasets,
and because it sheds light on the roles of linked selection and chromosomal inversions in 
establishing modern genetic diversity.
% Our results are consistent with what is known about linked selection and chromosomal inversions
% (a possible suprise being no evidence for segregating inversions in \textit{Medicago truncatula});
% the strength of the biological results lie in the unified framework that allows 
% flexible discovery and comparison between the three species.

Potential reviewers for this paper could include:
John Novembre (U Chicago, \href{mailto:jnovembre@uchicago.edu}{jnovembre@uchicago.edu}),
Joe Pickrell (NY Genome Center, \href{mailto:jkpickrell@nygenome.org}{jkpickrell@nygenome.org}),
David Lowry (MSU, \href{mailto:dlowry@msu.edu}{dlowry@msu.edu}),
or Andy Clark (Cornell, \href{mailto:ac347@cornell.edu}{ac347@cornell.edu}).

>>>>

We are writing to submit our paper,
``Local PCA Shows How the Effect of Population Structure Differs Along the Genome'',
for consideration in \textit{JOURNAL XXX}.
In this paper, we set out to describe how background patterns of relatedness 
(colloquially, ``population structure'')
vary along the genome at a scale larger than correlations induced by linkage.
It was not certain that we would find significant large-scale heterogeneity in patterns of relatedness;
in the field we tend to think of neutral, background population structure as a single thing,
uniform across the genome except at a few isolated loci under selection.
What we found was not only substantial variation at chromosomal scales,
but different underlying causes in the three species we consider.
The focus of our paper is on the method we developed to identify and visualize such variation in patterns of relatedness.
We believe our method represents a fundamental advance because it lets the data speak for itself,
providing an unbiased view of what are the most important variation in patterns of relatedness,
rather than looking for specific signals.
The method is therefore complementary to more common methods that identify small regions of the genome
that are outliers for particular summaries of relatedness (e.g., $F_{ST}$-based scans for selection).
We think the paper will be of interest to readers of \textit{JOURNAL XXX}
because the method fills an important gap in the genomicist's toolkit,
that is tailored to describing variation in modern, spatially sampled datasets,
and because it sheds light on the roles of linked selection and chromosomal inversions in 
establishing modern genetic diversity.
% Our results are consistent with what is known about linked selection and chromosomal inversions
% (a possible suprise being no evidence for segregating inversions in \textit{Medicago truncatula});
% the strength of the biological results lie in the unified framework that allows 
% flexible discovery and comparison between the three species.

Potential reviewers for this paper could include:
John Novembre (U Chicago, \href{mailto:jnovembre@uchicago.edu}{jnovembre@uchicago.edu}),
Joe Pickrell (NY Genome Center, \href{mailto:jkpickrell@nygenome.org}{jkpickrell@nygenome.org}),
Alkes Price (Harvard, \href{mailto:aprice@hsph.harvard.edu}{aprice@hsph.harvard.edu}),
or Andy Clark (Cornell, \href{mailto:ac347@cornell.edu}{ac347@cornell.edu}).
